---
title: "Final Project Rough Draft"
author:
  - "Darien Rodrigo"
  - "Rachel Le"
  - "Patrick Zhou"
format: html
embed-resources: true
---

# **Spotify Wrapped 2026.5**

```{r}
#| label: Load necessary packages
library(tidyverse)
```

## Part 1: Loading and Looking at Our Data

```{r}
spotify <- read_csv('https://raw.githubusercontent.com/FisherFraley/Datafest_Workshop/refs/heads/main/spotify_dataset.csv')
```

### Spotify Data Definitions

| **Column** | **Type** | **What does this variable mean?** |
|--------------------------|------------------|----------------------------|
| `...1` | Numeric | Row number |
| `track_id` | Character | Unique identifier for each track |
| `artist` | Character | Artist name(s) that is associated with the song. |
| `album_name` | Character | Name of the album that the song is in. |
| `song_name` | Character | Title of the song. |
| `popularity` | Numeric | Popularity score that ranges from 0-100, higher values indicate a greater popularity. |
| `duration_ms` | Numeric | Duration of the song in milliseconds. |
| `explicit` | Logical | Indicates if the song contains explicit content or not (`TRUE` or `FALSE`). |
| `danceability` | Numeric | A measure from 0-1 for how suitable the song is for dancing. A higher value indicates greater danceability. |
| `energy` | Numeric | A measure for how energetic the songs are from a scale of 0-1. Higher values indicate more intensity and activity. |
| `key` | Numeric | Musical key encoded from 0-11 it refers to its musical scale and harmonic root. |
| `loudness` | Numeric | Loudness of the track measured in decibels(dB). Values closer to 0 means that they are louder. |
| `mode` | Numeric | Music modality, 0 =minor, 1 = major |
| `speechiness` | Numeric | Speechiness measures the scale of present spoken words, scale of 0-1, values closer to one means there is more speech-like content. |
| `acousticness` | Numeric | Measures acoustic from a scale of 0-1. |
| `instrumentalness` | Numeric | A measure that indicates whether there is more vocal, less instrumental, or vice versa, scale of 0-1, values closer to 1, indicate more instrumental and less vocal. |
| `liveness` | Numeric | Measures presence of an audience in the recording, scales from 0-1, values closer to 1 suggests it is a live performance. |
| `valence` | Numeric | Measures the emotional tone of the track on a scale of 0-1, values closer to 1 indicate a happier or more positive sound. |
| `tempo` | Numeric | The tempo of the song measured by beats per minute (BPM) |
| `time_signature` | Numeric | The time measure of a song, it indicates the number of beats per measure it is on a scale of 0-5. |
| `track_genre` | Character | The genre assigned to the track. |
| `duration_min` | Numeric | Duration of the song in minutes. |
| `popularity_level` | Character | Popularity level, is the songs categorized from low, medium, high popularity, and it is derived from the popularity score. |

### Inspecting Our Data:

```{r}
glimpse(spotify)
head(spotify)
summary(spotify)
names(spotify)
```

What are the names of our columns?

What are some of the datatypes we're working on?

## Part 2: Cleaning our Data & Making it make more sense

Looking at our data, with all these artists and how much data we have on them, it would be a lot of trouble to make sure we have every piece of data that we need on each of these artists!

Luckily for us, we can use some handy functions to clean our data

First, we should check for missing values.

```{r}
colSums(is.na(spotify))
```

What columns have missing data?

Which columns have the most missing data?

Lets remove the missing data in the following steps:

1.  Make a fresh dataset & get rid of the missing entries

```{r}
spotify_clean <- spotify |>
  drop_na()
```

2.  Make Friendlier names

```{r}
spotify_renamed <- spotify_clean |>
  rename(
    song_name = track_name,
    artist = artists
  )
```

3.Make the song duration into minutes & add a better way to see popularity

```{r}
spotify_final <- spotify_renamed |>
  mutate(
    duration_min = duration_ms / 60000,
    popularity_level = case_when(
      popularity >= 80 ~ "High",
      popularity >= 50 ~ "Medium",
      TRUE ~ "Low"
    )
  )
```

## Part 3: Using Dplyr Verbs!

We've now taken a look at the dataset, cleaned it, made it make more sense, what is there to do now?

Obviously...

Work with our data! Why clean it and not use it?

In this section, we'll introduce some dplyr verbs that you'll come across.

Stop 1: Filter

Filter works as you'd imagine, it filters the results down

```{r}
spotify_final |>
  filter(popularity > 100)
```

Stop 2: Select You can select specific columns of a dataset suing select.

We'll make a smaller dataset consisting of core values

```{r}
spotify_small <- spotify_final|>
  select(artist, song_name, popularity)
```

Stop 3: Arrange

You can arrange the rows of a dataset by the observations within a column! Here, we'll arrange the top 10 by descending popularity.

```{r}
spotify_final |>
  arrange(desc(popularity)) |>
  head(10)
```

Stop 4: Summarize

```{r}
spotify_final |>
  summarize(
    avg_popularity = mean(popularity),
    avg_duration = mean(duration_min)
  )
```

Stop 5: Group Summaries

```{r}
spotify_final |>
  group_by(track_genre) |>
  summarize(avg_popularity = mean(popularity)) |>
  arrange(desc(avg_popularity))
```
