---
title: "Final Project Rough Draft"
author:
  - "Darien Rodrigo"
  - "Rachel Le"
  - "Patrick Zhou"
format: html
embed-resources: true
---

```{r}
#| label: setup
library(tidyverse)
```

## Part 1: Loading and Looking at Our Data

```{r}
spotify <- read_csv('https://raw.githubusercontent.com/FisherFraley/Datafest_Workshop/refs/heads/main/spotify_dataset.csv')
```

# Inspecting Our Data:

```{r}
glimpse(spotify)
head(spotify)
summary(spotify)
names(spotify)
```

What are the names of our columns?

What are some of the datatypes we're working on?

## Part 2: Cleaning our Data & Making it make more sense

Looking at our data, with all these artists and how much data we have on them, it would be a lot of trouble to make sure we have every piece of data that we need on each of these artists!

Luckily for us, we can use some handy functions to clean our data

First, we should check for missing values.

```{r}
colSums(is.na(spotify))
```

What columns have missing data?

Which columns have the most missing data?

Lets remove the missing data in the following steps:

1.  Make a fresh dataset & get rid of the missing entries

```{r}
spotify_clean <- spotify |>
  drop_na()
```

2.  Make Friendlier names

```{r}
spotify_renamed <- spotify_clean |>
  rename(
    song_name = track_name,
    artist = artists
  )
```

3.Make the song duration into minutes & add a better way to see popularity

```{r}
spotify_final <- spotify_renamed |>
  mutate(
    duration_min = duration_ms / 60000,
    popularity_level = case_when(
      popularity >= 80 ~ "High",
      popularity >= 50 ~ "Medium",
      TRUE ~ "Low"
    )
  )
```

## Part 3: Using Dplyr Verbs!

We've now taken a look at the dataset, cleaned it, made it make more sense, what is there to do now?

Obviously...

Work with our data! Why clean it and not use it?

In this section, we'll introduce some dplyr verbs that you'll come across.

Stop 1: Filter

Filter works as you'd imagine, it filters the results down

```{r}
spotify_final |>
  filter(popularity > 100)
```

Stop 2: Select You can select specific columns of a dataset suing select.

We'll make a smaller dataset consisting of core values

```{r}
spotify_small <- spotify_final|>
  select(artist, song_name, popularity)
```

Stop 3: Arrange

You can arrange the rows of a dataset by the observations within a column! Here, we'll arrange the top 10 by descending popularity.

```{r}
spotify_final |>
  arrange(desc(popularity)) |>
  head(10)
```

Stop 4: Summarize

```{r}
spotify_final |>
  summarize(
    avg_popularity = mean(popularity),
    avg_duration = mean(duration_min)
  )
```

Stop 5: Group Summaries

```{r}
spotify_final |>
  group_by(track_genre) |>
  summarize(avg_popularity = mean(popularity)) |>
  arrange(desc(avg_popularity))
```
