---
title: "Final Project Rough Draft"
author:
  - "Darien Rodrigo"
  - "Rachel Le"
  - "Patrick Zhou"
format: html
embed-resources: true
---

# **Spotify Wrapped 2026.5**

Have you ever wondered how your Spotify Wrapped comes together? Every year, Spotify users get a snapshot of their listening habits, including how many minutes they've spent on Spotify listening to songs, their most streamed song, and their top artist. Data scientists at Spotify are asked to develop algorithms to parse through copious amounts of user data and then present them in digestible and often visually intriguing ways. Now that is what we will do! What better time than the present to make our own Spotify Wrapped now that 2026 is halfway through! In the next several questions, you will be asked to clean, compile, and then visualize some of the top artists from this dataset.

```{r}
#| label: Load necessary packages
library(tidyverse)
library(knitr)
library(purrr)
```

## Part 1: Loading and Looking at Our Data

```{r}
spotify <- read_csv('https://raw.githubusercontent.com/FisherFraley/Datafest_Workshop/refs/heads/main/spotify_dataset.csv')
```

Our dataset is quite intimidating! Breaking it down and understanding what constitutes it makes it easier to understand what we *do* and do *not* need.

### Spotify Data Definitions

Below are the meanings of what each column represents—take note of which are numeric, character, or logical variables.

| **Column** | **Type** | **What does this variable mean?** |
|------------------------|------------------------|------------------------|
| `...1` | Numeric | Row number |
| `track_id` | Character | Unique identifier for each track |
| `artist` | Character | Artist name(s) that is associated with the song. |
| `album_name` | Character | Name of the album that the song is in. |
| `song_name` | Character | Title of the song. |
| `popularity` | Numeric | Popularity score that ranges from 0-100, higher values indicate a greater popularity. |
| `duration_ms` | Numeric | Duration of the song in milliseconds. |
| `explicit` | Logical | Indicates if the song contains explicit content or not (`TRUE` or `FALSE`). |
| `danceability` | Numeric | A measure from 0-1 for how suitable the song is for dancing. A higher value indicates greater danceability. |
| `energy` | Numeric | A measure for how energetic the songs are from a scale of 0-1. Higher values indicate more intensity and activity. |
| `key` | Numeric | Musical key encoded from 0-11 it refers to its musical scale and harmonic root. |
| `loudness` | Numeric | Loudness of the track measured in decibels(dB). Values closer to 0 means that they are louder. |
| `mode` | Numeric | Music modality, 0 =minor, 1 = major |
| `speechiness` | Numeric | Speechiness measures the scale of present spoken words, scale of 0-1, values closer to one means there is more speech-like content. |
| `acousticness` | Numeric | Measures acoustic from a scale of 0-1. |
| `instrumentalness` | Numeric | A measure that indicates whether there is more vocal, less instrumental, or vice versa, scale of 0-1, values closer to 1, indicate more instrumental and less vocal. |
| `liveness` | Numeric | Measures presence of an audience in the recording, scales from 0-1, values closer to 1 suggests it is a live performance. |
| `valence` | Numeric | Measures the emotional tone of the track on a scale of 0-1, values closer to 1 indicate a happier or more positive sound. |
| `tempo` | Numeric | The tempo of the song measured by beats per minute (BPM) |
| `time_signature` | Numeric | The time measure of a song, it indicates the number of beats per measure it is on a scale of 0-5. |
| `track_genre` | Character | The genre assigned to the track. |
| `duration_min` | Numeric | Duration of the song in minutes. |
| `popularity_level` | Character | Popularity level, is the songs categorized from low, medium, high popularity, and it is derived from the popularity score. |

### Inspecting Our Data

1.  What are the names of our columns?

```{r}
names(spotify)
```

2.  What are some of the datatypes we're working with?

```{r}
glimpse(spotify)
```

## Part 2: Clean the Data!

Looking at our data, with all these artists and how much data we have on them, it would be a lot of trouble to make sure we have every piece of data that we need on each of these artists!

Luckily for us, we can use some handy functions to clean our data. Create a new dataframe named `spotify_cleaned` and perform the following tasks:

::: {.callout-note}
Before analyzing the Spotify data, make sure you remove rows with missing values in important columns
:::

3.  First, check for missing values.

```{r}
colSums(is.na(spotify))
```

4.  What columns have missing data?


```{r}
# Track Name, Artists, Album Name have missing data
```

5.  There are 21 total columns in our dataset. Remove all columns *except* the following: `track_id`, `artists`, `album_name`,`track_name`, `popularity`, `duration_ms`, and `track_genre`.

```{r}
spotify_clean <- spotify |>
  select(track_id, artists, album_name, track_name, popularity, duration_ms, track_genre)
```

6.  Which rows have important information missing? In other words, if rows are missing `track_name`, `artists`, or `album_name` should they even be included in our data set? Remove rows where any of the aforementioned columns contain `NA`.

```{r}
spotify_clean <- na.omit(spotify_clean)
```

7.  Are there any duplicate rows? Remove them!

```{r}
spotify_clean <- unique(spotify_clean)
```

8.  Note that column `duration_ms` is in milliseconds. It's quite tough to imagine a song in terms of milliseconds. Convert these milliseconds to minutes.

```{r}
spotify_clean <- spotify_clean |>
  mutate(
    duration_min = duration_ms / 60000
  )
```




## Part 3: Use the Clean Dataset for Analysis

9. Now that your dataset has been cleaned, you can begin exploring the songs that make up Spotify's catalog. Before identifying trends, it is useful to understand what a "typical" song looks like.

    Calculate summary statistics for the following variables across all songs:

    -   Popularity

    -   Duration (in minutes)

    -   Danceability

    -   Energy

    Based on these summaries, describe the average song in this dataset. What characteristics seem most common?

```{r}
spotify <- spotify |>
  mutate(
    duration_min = duration_ms / 60000
  )
summary(
  spotify[, c("popularity","duration_min","danceability","energy")]
)
```

10. Spotify assigns every song a popularity score ranging from 0 to 100. Data scientists often begin by examining how a variable is distributed before drawing conclusions from it.

    Create an appropriate visualization of song popularity.

    After creating your visualization, answer the following:

    -   Is song popularity normally distributed?

    -   Are most songs highly popular or relatively unknown?

    -   Are there any unusual patterns or outliers that stand out?

```{r}
ggplot(spotify_clean,
       aes(x = popularity)) +
  geom_histogram(binwidth = 3,
                 fill = "darkgreen",
                 color = "black") +
  labs(
    title = "Distribution of Song Popularity",
    x = "Popularity",
    y = "Score"
  )
```
Song Popularity looks to be mostly normally distributed. The curve sits a little left of 50 so that means more songs are lesser known. There is a large outlier sitting on 0.


11. Imagine that this dataset represents your listening history for the first half of 2026. Spotify Wrapped often highlights a listener's favorite artists and genres.

    Identify:

    -   Your top artist (the artist appearing most often)

    -   Your top five artists

    -   Your top genre

    Create a table summarizing your findings.

    What do these results suggest about your listening habits?

```{r}
top_artist <- spotify_clean |>
  arrange(desc(popularity)) |>
  select(artists)
head(top_artist, 1)


head(top_artist)

top_genre <- spotify_clean |>
  arrange(desc(popularity)) |>
  select(track_genre)

head(top_genre, 1)
```

12. Suppose Spotify generates a random playlist containing 25 songs from its catalog.

    Would that playlist tend to contain highly popular songs, or would it mostly contain average songs?

    To investigate:

    1.  Randomly sample 25 songs.

    2.  Compute the playlist's average popularity.

    3.  Repeat this process 1,000 times.

    Visualize the resulting distribution of average popularity scores.

    Based on your simulation, what popularity score would you expect for a typical playlist?

::: {.callout-tip}
For the playlist simulation, you can use `map_dbl()` to repeat the random sampling process many times.
:::

```{r}
set.seed(100)

rand_playlist <- spotify_clean |>
  map_dbl(
  .x = 1:1000,
  .f = ~ mean(sample(spotify_clean$popularity, size = 25))
)


hist(
  rand_playlist,
  main = "Distribution of Average Playlist Popularity",
  xlab = "Average Popularity",
  col = "steelblue",
  border = "white"
)

mean(rand_playlist)

sd(rand_playlist)
```
